---
title: "Appendix B: In-Person Students"
output: word_document
---

```{r IPinit, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE,error=FALSE,fig.width=9,fig.height=9,
  cache=TRUE)
options(knitr.kable.NA='-')
```

```{r IPloadCode, include=FALSE}
library(tidyverse)
library(RItools)
library(lmerTest)
library(lme4)
library(texreg)

source("../code/functions.r")
Scale <- function(x) (x-mean(x,na.rm=TRUE))/sd(x,na.rm=TRUE)
condNames=c(FH2T='FH2T',Dragon='DragonBox',ASSISTments='Immediate Feedback',BAU='active Control')
```

```{r IPloadData}
## randomized data
datIP0 <- read_csv("../data/DATA20220202_4092.csv",na = c("", "NA","#NULL!"))%>%
  filter(
    !endsWith(rdm_condition,'Resource'),
    virtual==0
    )%>%
  mutate(
    ScaleScore7 = Scale(Scale.Score7),
    hasPretest=is.finite(pre.total_math_score),
    hasPosttest=is.finite(post.total_math_score),
    Z =rdm_condition,
    race=raceEthnicityFed%>%
               factor()%>%
               fct_other(keep=c("1","3","6"))%>%
               fct_recode(Hispanic="1",Asian="3",White="6")%>%
            fct_relevel("White"),
      ScaleScore5 = Scale(`Scale.Score5`),
      accelerated=grepl('Accelerated',courseName),
      pretest = pre.total_math_score-round(mean(pre.total_math_score,na.rm=TRUE))
  )%>%
  filter(!is.na(Z))%>%
  mutate(
    teach=ifelse(is.na(TeaIDPre),TeaIDEnd,TeaIDPre),
    class=ifelse(is.na(ClaIDPre),ClaIDEnd,ClaIDPre))

noPre=datIP0%>%
  group_by(SchIDPre)%>%
  summarize(pretest=mean(hasPretest))%>%
  filter(pretest<0.01)%>%
  pull(SchIDPre)
### Which teachers?
noPreTch=unique(datIP0$TeaIDPre[datIP0$SchIDPre==noPre])

datIP0$dropsch1a=datIP0$teach%in%noPreTch

diffStuds=datIP0$StuID[datIP0$DROPSCH1==1 & !datIP0$dropsch1a]
ndiff=length(diffStuds)

datIP0$Z <- relevel(factor(datIP0$Z),ref='BAU')

covNames <-
    c("pretest","race","FEMALE","EIP","IEP","GIFTED","accelerated") ## more?
```


```{r IPanalysisSample}
datIP0 <- datIP0%>%filter(!dropsch1a)


### analysis sample
datIP <- datIP0%>%
    filter(DROPSCH2==0,hasPretest,hasPosttest)
```

Both overall and differential attrition was substantially worse for students who began the year with remote instruction, as opposed to in-person instruction.
This section repeats the attrition analysis and the main HLM analysis for the subset of students who started the academic year in-person.

## Attrition Analysis

### Overall and Differential Attrition Rates for In-Person Students
```{r IPwccAttritionNumbers, fig.cap="Figure B.1: Overall and Differential Attrition for In-Person Students, Plotted Against WWC Standards."}

Ncond=table(datIP0$Z)
N=sum(Ncond)
ncond=table(datIP$Z) #
n=sum(ncond)

attCond=1-ncond/Ncond

diffAtt=outer(attCond,attCond,"-")
overallPair=diffAtt
for(cc in names(ncond))
  for(dd in names(ncond))
    overallPair[cc,dd]=1-(ncond[cc]+ncond[dd])/(Ncond[cc]+Ncond[dd])

tab=rbind(
  `# Randomized`=Ncond,
  `# in Analysis Sample`=ncond,
  `# Attrition Rate (%)`=attCond*100)
colnames(tab)=condNames[colnames(tab)]
tab=cbind(Overall=c(N,n,100*(1-n/N)),tab)
kable(tab,digits=1,caption="Table B.1: Attrition rates by experimental condition")

wwc <- read.csv("../attritionReports/wwc.csv")
names(wwc)[1] <- "Overall"


with(wwc,
  plot(Overall,Differential1,type="l",ylim=c(0,11),
       main="WWC Attrition Standards",
       xlab="Overall Attrition (In-Person)", ylab="Differential Attrition (In-Person)"))
  polygon(c(0,0,65,65),c(0,11,11,0),col="red")
  polygon(c(0,wwc[[1]],65),c(0,wwc$Differential1,0),col="yellow")
  polygon(c(0,wwc[[1]]),c(0,wwc$Differential0),col="green")

diffAtt=abs(diffAtt)

for(i in 1:3)
  for(j in (i+1):4){
    if('BAU'%in%colnames(diffAtt)[c(i,j)]){
      ov=overallPair[i,j]*100
      diff=diffAtt[i,j]*100
      points(ov,diff,pch=16)
    }
    # text(ov,diff,
    #      paste(condNames[i],'vs',condNames[j]),pos=2)
  }

text(
  overallPair['Dragon','BAU']*100,
  diffAtt['Dragon','BAU']*100,
  paste(condNames['Dragon'],'vs.',condNames['BAU']),
  pos=1)
text(
  overallPair['ASSISTments','BAU']*100,
  diffAtt['ASSISTments','BAU']*100,
  paste(condNames['ASSISTments'],'vs.',condNames['BAU']),
  pos=2)
text(
  overallPair['FH2T','BAU']*100,
  diffAtt['FH2T','BAU']*100,
  paste(condNames['FH2T'],'vs.',condNames['BAU']),
  pos=2)
```

### Baseline Equivalence for In-Person Students


Figure B.2 plots covariate distributions across the four treatment groups; little imbalance is apparent.



```{r IPbalancePlot, fig.cap="Figure B.2: Comparing covariates across conditions"}
### balance plot:
datIP%>%
    mutate(race=substr(as.character(race),1,1))%>%
  balPlot(covNames,trtVar="Z",data=.)
```




```{r IPeffectSizes}
##### standardized differences following WWC procedures 4.1 chp VI p. 15
dCox=function(x,z){
  pi=mean(x[z],na.rm=TRUE)
  pc=mean(x[!z],na.rm=TRUE)
  (qlogis(pi)-qlogis(pc))/1.65
}

hedgesG=function(x,z){
  yi=mean(x[z],na.rm=TRUE)
  yc=mean(x[!z],na.rm=TRUE)
  ni=sum(!is.na(x[z]))
  nc=sum(!is.na(x[!z]))
  N=ni+nc
  omega=1-3/(4*N-9)
  omega*(yi-yc)/sqrt(((ni-1)*var(x[z])+(nc-1)*var(x[!z]))/(N-2))
}

stdDiffs=list()
for(cond in levels(datIP$Z)[-1]){
  diffDat=
    datIP%>%
    filter(Z%in%c(cond,'BAU'))%>%
    mutate(Z=Z==cond)%>%
    select(all_of(covNames),Z)
  diffs=map_dbl(setdiff(covNames,c('pretest','race'))%>%setNames(.,.),
                ~dCox(diffDat[[.]],diffDat$Z))
  for(r in levels(diffDat$race)[-1]){
    diffs=c(diffs,dCox(diffDat$race==r,diffDat$Z))
    names(diffs)[length(diffs)] <- r
  }
  diffs=c(diffs,pretest=hedgesG(diffDat$pretest,diffDat$Z))
  stdDiffs[[cond]] <- diffs
}
diffs <- as.data.frame(stdDiffs)

lapply(diffs,
       function(x)
         paste0(round(x,3),
                ifelse(abs(x)>0.05,'*','')))%>%
  as.data.frame(row.names=rownames(diffs))%>%
  kable(caption="Table B.2: Effect sizes for covariate mean differences between each of the experimental conditions and the Active Control for in-person students. Stars indicate effect sizes >0.05, for which WWC recommends statistical adjustment",
        digits=3)
```

```{r IPritools}
covForm <- as.formula(paste("~",paste(covNames,collapse="+")))
totForm <- update(covForm,Z~.)
bals <- xbalMult(totForm,datIP,trtLevs=unique(datIP$Z),strata=list(noBlocks=NULL,cls=~class),na.rm=TRUE)

overall <- map_dfr(grep('BAU',names(bals),value=TRUE),~c(contrast=.,bals[[.]]$overall['cls',]))
nnn=gsub(' |vs\\.|BAU','',overall$contrast)
overall$contrast=c(ASSISTments='Immediate Feedback',Dragon='DragonBox',FH2T="FH2T")[nnn]

kable(overall,caption='Table B.3: Overall covariate balance for each of the Active Control comparisons for In-Person students',digits=3)
```

```{r IPindividualBalance}
ind <- map(bals[grep('BAU',names(bals),value=TRUE)],~data.frame(.$results[,c('z','p'),'cls']))
ind <- map(ind,~data.frame(Z=paste0(-round(.$z,3),ifelse(.$p<0.001,'***',ifelse(.$p<0.01,'**',ifelse(.$p<0.05,'*',ifelse(.$p<0.1,'.',''))))),row.names = rownames(.)))
nnn=gsub(' |vs\\.|BAU','',names(ind))
ind=as.data.frame(ind,row.names = rownames(ind[[1]]))
names(ind)=c(ASSISTments='Immediate Feedback',Dragon='DragonBox',FH2T="FH2T")[nnn]
kable(ind,caption='Table B.4: Z-scores testing baseline covariate balance for in-person students between each experimental condition and the Active Control')
```

## HLM Models

```{r IPhlm,results='asis'}
datIP$post=datIP$post.total_math_score
datIP$pre=datIP$pre.total_math_score

datIP$ncomp=rowSums(datIP[,startsWith(names(datIP),'complete_assignment')])
datIP$percomp=datIP$ncomp/9
#datIP$percomp=datIP$percomp-mean(datIP$percomp)

datIP=mutate(datIP,
             FH2T=ifelse(Z=='FH2T',1,0),
             DragonBox=ifelse(Z=='Dragon',1,0),
             Instant=ifelse(Z=='ASSISTments',1,0),
             Asian=ifelse(race=='Asian',1,0),
             Hispanic=ifelse(race=='Hispanic',1,0),
             Other=ifelse(race=='Other',1,0))

#datIP$pre=datIP$pre-mean(datIP$pre)
center=function(x) scale(x,center=TRUE,scale=FALSE)
mod1=lmer(post~FH2T+DragonBox+Instant+(1|class)+(1|teach),data=datIP)
mod2=update(mod1,.~.+center(pre)+Asian+Hispanic+Other+MALE+GIFTED+accelerated+EIP+IEP)
mod3=update(mod2,.~.+center(percomp))
mod4=update(mod3,.~.+center(pre):(FH2T+DragonBox+Instant))


## list(mod1,mod2,mod3,mod4)%>%
##   htmlreg(#file="tableZZ4orig.html",
##     ci.force=TRUE,single.row=TRUE,
##     reorder.coef=c(1,4,3,2,5:14,17,16,15),digits=3,
##     doctype = FALSE)#,star.symbol="\\.")


stars=function(p)
  ifelse(p<0.001,'***',
  ifelse(p<0.01,'**',
  ifelse(p<0.05,'*',
         ifelse(p<0.1,'.',''))))

modSumm=function(mod){
  ccc=summary(mod)$coef
  data.frame(
    x=rownames(ccc),
    coef=paste0(round(ccc[,1],2),stars(ccc[,5])),
    ci=confint(mod,method='Wald')[-c(1:3),]%>%round(2)%>%apply(1,function(x) paste0('[',x[1],',',x[2],']'))
  )
}

map(list(mod1,mod2,mod3,mod4),
    modSumm)%>%
  reduce(full_join,by='x')%>%
  kable(col.names=c(' ','Model 1', ' ', 'Model 2', ' ','Model 3',' ','Model 4',' '),
        caption='HLMs for in-person students')

```
