---
title: "Attrition Analysis: Delayed Assessment, Including Pretest No-Shows"
output: word_document
---

```{r init, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE,error=FALSE,fig.width=6.5,fig.height=7)
```

```{r}
library(tidyverse)
library(RItools)
library(mosaic)
source('../code/functions.r')

dat <- read_csv('../data/Assessment_merged_2021_06_16_state_assessment_N=4311.csv')

### attrition variables
dat <- dat%>%
    mutate(
        hasPretest=is.finite(pre_MA_total_score),
        hasMidtest=is.finite(mid_MA_total_score),
        hasPosttest=is.finite(post_MA_total_score),
        hasDelayed=is.finite(delayed.total_math_score)
        )
### what's the difference between hasDelayed & complete_delayed_posttest?

dat$Z <- dat$rdm_condition ## differentiates between * and *-Resource (what does this mean?)
```

```{r}
Scale <- function(x) (x-mean(x,na.rm=TRUE))/sd(x,na.rm=TRUE)

dat <- dat%>%
    mutate(
        race=student_raceEthnicityFed%>%
               factor()%>%
               fct_lump_min(200)%>%
               fct_recode(`Hispanic/Latino`="1",Asian="3",White="6",Black="4")%>%
            fct_relevel('White'),
        ScaleScore7 = Scale(Scale.Score_7th.grade),
        ScaleScore5 = Scale(`ScaleScore_5th grade`),
        pretest = pre.total_math_score-round(mean(pre.total_math_score,na.rm=TRUE))#Scale(pre.total_math_score),
        )

covNames <-
    c('pretest','ScaleScore5','ScaleScore7','race','FEMALE','EIP','ESOL','IEP','GIFTED') ## more?
```

Attrition by school:
```{r}
## a couple schools basically just dropped out
dat%>%
    group_by(initial_school_id)%>%
    summarize(n=n(),across(starts_with('has'),mean))%>%
    kable(digits=3)
```

Drop schools with <0.01 posttest scores (S03 & S07)
```{r}

dat$initial_school_id[is.na(dat$initial_school_id)] <- 'SNA'

## drop schools where (almost) everyone dropped out (S03 & S07)
dat <- dat%>%
    group_by(initial_school_id)%>%
    mutate(postPer=mean(hasPosttest))%>%
    filter(postPer>0.01)%>%
    ungroup()%>%
    select(-postPer)

```

Attrition rates by condition:
```{r attritionRates}
dat%>%
    bind_rows(mutate(dat,Z='Overall'))%>%
    mutate(Z=factor(Z,levels=c('Overall',unique(dat$Z))))%>%
    group_by(Z)%>%
    summarize(n_assigned=n(),across(starts_with('has'),mean))%>%
    kable(digits=3)
```

Excluding students with missing pre-test:

```{r }
dat%>%
    bind_rows(mutate(dat,Z='Overall'))%>%
    filter(hasPretest)%>%
    mutate(Z=factor(Z,levels=c('Overall',unique(dat$Z))))%>%
    group_by(Z)%>%
    summarize(n_assigned=n(),across(starts_with('has'),mean))%>%
    kable(digits=3)
```


### between randomization & pretest
differential attrition:
```{r}
att <- dat%>%
    group_by(Z)%>%
    summarize(att=1-mean(hasPretest))%>%
    ungroup()
diff <- outer(att$att,att$att,FUN='-')
rownames(diff) <- colnames(diff) <- att$Z
kable(diff,digits=3)
```

### Delayed
#### Including people w/o pretests:
differential attrition:
```{r}
att <- dat%>%
    group_by(Z)%>%
    summarize(att=1-mean(hasDelayed))%>%
    ungroup()
diff <- outer(att$att,att$att,FUN='-')
rownames(diff) <- colnames(diff) <- att$Z
kable(diff,digits=2,caption='mid-test differential attrition, including pre-test no-shows')
```


Balance plot:
```{r}
dat%>%
    filter(hasDelayed)%>%
    mutate(race=substr(as.character(race),1,1))%>%
    balPlot(covNames,trtVar='Z',data=.)
```

Balance tests:
```{r}
ddd <- dat%>%
    filter(hasDelayed,!is.na(Z), !endsWith(Z,"Resource"))

tests <- map_dfr(covNames,~balTestOne(ddd[[.]],ddd$Z,ddd$initial_teacher_class))
tests <- tests[,sapply(tests,function(x) all(!is.na(x)))]
tests$p.adj <- p.adjust(tests$p.value,method='fdr')
tests$covariate <- covNames
tests%>%
    select(covariate,method,statistic,p.value,p.adj)%>%
    kable()
```
xBalance version:
```{r}
covForm <- as.formula(paste('~',paste(covNames,collapse='+')))
totForm <- update(covForm,Z~.)

bals <- xbalMult(totForm,ddd,trtLevs=unique(ddd$Z)[!endsWith(unique(ddd$Z),'Resource')],strata=list(cls=~ddd$initial_teacher_class),na.rm=TRUE)
plotXbals(bals)

map_dfr(names(bals),~tibble(Comparison=.,Overall.p=bals[[.]]$overall[1,'p.value']))%>%
    kable(digits=3)
```
